---
title: "Pre-announcing mizer 3.1"
description: >
  An experimental second-order accurate numerical scheme in size, a new
  L-stable time-stepper, and a long-overdue clarification of the maximum-size
  parameters.
author: "Gustav Delius"
date: "2026-06-24"
categories: [release]
---

We are pleased to announce the release of **mizer 3.1**. Coming so soon after the big 3.0 release, this is a more focused update: its centrepiece is an *experimental, opt-in* numerical scheme that makes mizer second-order accurate in the size variable, backed up by a new time-stepping method. Alongside the numerics there is a clarification of how mizer handles the various "maximum size" parameters, plus the usual collection of smaller improvements and bug fixes.

If you do nothing differently, **your results will not change**: the new accuracy features are switched off by default and the default code path is byte-for-byte identical to mizer 3.0.

We will release this new version on CRAN in a few days. We welcome bug reports and feature requests on [GitHub](https://github.com/sizespectrum/mizer/issues).

## Experimental: second-order accuracy in size

mizer solves the size-spectrum (McKendrick–von Foerster) equation on a logarithmic grid of size bins using a finite-volume method. Historically that method has been *first order* in the bin width: halving the bin width roughly halves the discretisation error. That is robust and fast, but on the coarse grids people often use it can introduce a visible bias — most noticeably a small shift in size-integrated quantities such as biomass, yield and SSB.

mizer 3.1.0 introduces an experimental **second-order** finite-volume scheme. The idea is to stop point-sampling quantities at the left edge of each size bin and instead use their exact (or trapezoidal) average *over* the bin, so that sinks, sources, fluxes and diagnostics are all consistent with the finite-volume interpretation of the abundance as a cell average. With it switched on, halving the bin width quarters the error.

It is controlled by a single new slot, accessed with `second_order_w()`:

``` r
# Turn on the full second-order scheme
second_order_w(params) <- TRUE

# Or control the two ingredients independently
second_order_w(params) <- c(flux = "centred", bin_average = TRUE)
```

There are two ingredients, and a fully second-order run wants both:

- **`bin_average`** replaces every point-sampled power law and quadrature that feeds the update — external mortality and diffusion, the resource growth rate and capacity, the predation kernels, gear selectivity and the reproductive investment — with its exact bin average. The summary integrals (`getBiomass()`, `getSSB()`, `getYield()`, …) and the size-resolved mortalities are likewise weighted and placed consistently, the latter now reported at the geometric bin centre where a bin average actually lives.
- **`flux`** chooses the advective reconstruction for the growth term: `"van_leer"` is a TVD scheme that stays second order while keeping abundances non-negative, and `"centred"` is genuinely second order even at peaks and troughs.

You can also build a model with the scheme already enabled by passing `second_order_w = TRUE` to `newMultispeciesParams()`, `newTraitParams()`, `newCommunityParams()` or `newSingleSpeciesParams()`.

### Does it actually deliver?

It is one thing to claim second-order accuracy and another to measure it. mizer ships with an [analytic test vignette](https://sizespectrum.org/mizer/articles/analytic_test.html) that solves the transport equation exactly for power-law rates, so we can compare the numerical solution against a known answer. The figure below — adapted from that vignette — projects a travelling, diffusing pulse on grids of increasing resolution and plots the relative error against the (logarithmic) grid spacing `Δx`, once with the default first-order scheme and once with `second_order_w <- TRUE`.

```{r}
#| label: spatial-convergence
#| message: false
#| code-fold: true
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Relative L2 error of a projected pulse versus grid spacing. The default scheme converges roughly linearly (slope-1 guide line); the experimental second-order scheme follows slope 2 and is up to ~100x more accurate over the range of resolutions shown."
library(mizer)

# Power-law rates with an exact time-dependent (advecting, diffusing, decaying)
# solution. See the analytic_test vignette for the derivation.
p <- 0.7; A <- 1; B <- 0.5; K <- 0.1
w0 <- 1e-2; t_start <- 0.1; t_end <- 2

# Growth rate g(w) = A w^p, supplied as a custom rate function.
start_growth <- function(params, ...) matrix(A * params@w^p, nrow = 1)

# Exact pulse N(w, t): a Green's function of the transport equation.
N_analytic <- function(w, t, w0, t0, K_eff = K) {
    U <- A - 0.5 * K_eff; V <- 0.5 * K_eff * (1 - p); b <- B / (1 - p)
    nu <- sqrt((U / V)^2 + 4 * b / V)
    dt <- t - t0
    x  <- w^(1 - p) / (1 - p); x0 <- w0^(1 - p) / (1 - p)
    z  <- 2 * sqrt(x * x0) / (V * dt)
    log_N <- -log(V * dt) + (U / (2 * V)) * log(x / x0) -
             (x + x0) / (V * dt) + z + log(besselI(z, nu, expon.scaled = TRUE))
    exp(log_N) * w^(-p)
}

# Single-species model with these rates, optionally second order in size.
build_sp <- function(no_w, second_order) {
    sp <- data.frame(species = "Test", w_max = 1000, w_mat = 100,
                     n = p, z0 = 0, z_ext = B, d = p - 1, D_ext = K)
    pr <- newMultispeciesParams(sp, no_w = no_w, min_w = 1e-3, info_level = 0)
    second_order_w(pr) <- second_order
    pr <- setRateFunction(pr, "EGrowth", "start_growth")
    pr@interaction[] <- 0                      # mortality is purely B w^(p-1)
    pr <- setExtMort(pr)
    pr@species_params$constant_reproduction <- 0
    setRateFunction(pr, "RDD", "constantRDD")
}

# In the second-order scheme N_j is the cell *average* (≈ value at the cell
# centre); in the first-order scheme it is the node value.
ref_w <- function(pr, second_order) {
    wv <- w(pr)
    if (second_order) wv * sqrt(wv[2] / wv[1]) else wv
}

# Relative L2 error of the projected pulse, restricted to its bulk.
pulse_err <- function(no_w, second_order, dt = 0.0025) {
    pr <- build_sp(no_w, second_order)
    rw <- ref_w(pr, second_order)
    initialN(pr) <- matrix(N_analytic(rw, t_start, w0, 0), nrow = 1)
    sim <- project(pr, t_max = t_end - t_start, dt = dt,
                   t_save = t_end - t_start, method = "tr_bdf2",
                   progress_bar = FALSE)
    num <- finalN(sim)[1, ]; ana <- N_analytic(rw, t_end, w0, 0)
    mask <- ana > max(ana) * 1e-3
    sqrt(sum((num[mask] - ana[mask])^2 * pr@dw[mask]) /
         sum(ana[mask]^2 * pr@dw[mask]))
}

no_ws <- c(50, 100, 200, 400, 800)
dx <- log(1000 / 1e-3) / no_ws                 # log-size grid spacing
err_first  <- sapply(no_ws, pulse_err, second_order = FALSE)
err_second <- sapply(no_ws, pulse_err, second_order = TRUE)
sp_order <- function(e) unname(coef(lm(log(e) ~ log(dx)))[2])

plot(dx, err_first, log = "xy", type = "b", pch = 16, lwd = 2,
     col = "#E69F00", xlab = expression(Delta * x == log * beta),
     ylab = "relative L2 error", ylim = range(err_first, err_second),
     main = "Order of convergence in size")
lines(dx, err_second, type = "b", pch = 17, lwd = 2, col = "#0072B2")
i0 <- length(dx)
lines(dx, err_first[i0]  * (dx / dx[i0])^1, lty = 3, col = "grey50")
lines(dx, err_second[i0] * (dx / dx[i0])^2, lty = 2, col = "grey50")
legend("bottomright", bty = "n",
       legend = c(sprintf("first order (slope %.2f)", sp_order(err_first)),
                  sprintf("second order (slope %.2f)", sp_order(err_second)),
                  "slope 1", "slope 2"),
       col = c("#E69F00", "#0072B2", "grey50", "grey50"),
       pch = c(16, 17, NA, NA), lty = c(1, 1, 3, 2), lwd = 2)
```

The default scheme converges only linearly: its error falls roughly in step with the slope-1 guide line, so halving the grid spacing roughly halves the error (the coarsest grids here are still pre-asymptotic, which is why the fitted slope sits a little below 1). The experimental scheme follows the slope-2 line — about four times more accurate for every halving of `Δx` — so it pulls away fast: from a few times more accurate on the coarsest grid to roughly a hundredfold more accurate on the finest grid shown.

### What the error looks like: numerical diffusion

That abstract error norm has a very concrete consequence. A first-order upwind scheme does not merely lose accuracy — it injects **numerical diffusion**, an artificial smearing that imitates the individual growth variability mizer models with its diffusion term, except that here it is a numerical artefact rather than biology. We can watch it happen by following annual cohorts. Driving the model with a once-a-year pulse of reproduction (as in the [cohort dynamics vignette](https://sizespectrum.org/mizer/articles/cohort_dynamics_and_diffusion.html)) and switching physical diffusion **off**, every individual born together grows together, so each cohort should stay a sharp spike for life. Any broadening we see is the numerics talking, not the model.

```{r}
#| label: cohort-diffusion
#| message: false
#| code-fold: true
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Five annual cohorts grown with no physical diffusion, so any spreading is purely numerical. The default first-order upwind scheme smears the cohorts into low, broad, merging humps; the second-order scheme keeps each one tall and sharp."
# Pulsed annual reproduction: one sharp cohort per year.
pulse_width <- 0.1
annual_pulse_RDD <- function(rdi, species_params, t, ...) {
    frac <- t %% 1
    if (frac < pulse_width) {
        rdd <- rdi
        rdd[] <- 1 - cos(frac / pulse_width * 2 * pi)
        rdd
    } else 0 * rdi
}

run_cohorts <- function(second_order) {
    pr <- newSingleSpeciesParams(h = 10, no_w = 400)
    pr <- setRateFunction(pr, "RDD", "annual_pulse_RDD")
    initialN(pr)[] <- 0                  # start from an empty spectrum
    second_order_w(pr) <- second_order   # physical diffusion stays off
    project(pr, t_max = 5, dt = 0.005, t_save = 0.05,
            method = "tr_bdf2", progress_bar = FALSE)
}

sim_first  <- run_cohorts(FALSE)         # default first-order upwind flux
sim_second <- run_cohorts(TRUE)          # second-order scheme

wv <- w(sim_first@params)
b_first  <- finalN(sim_first)[1, ]  * wv^2
b_second <- finalN(sim_second)[1, ] * wv^2

plot(wv, b_second, type = "l", log = "x", lwd = 2, col = "#0072B2",
     xlim = c(1e-2, 1e2),
     xlab = "Weight [g]", ylab = expression(Biomass~density~~N(w) * w^2),
     main = "Numerical broadening of cohorts (no physical diffusion)")
lines(wv, b_first, lwd = 2, col = "#E69F00")
legend("topright", bty = "n", lwd = 2, col = c("#E69F00", "#0072B2"),
       legend = c("default (first-order upwind)", "second_order_w = TRUE"))
```

With the default upwind flux the cohorts bleed into low, broad humps and the oldest are already merging — even though nothing in the model is making them spread. The second-order scheme keeps each cohort roughly three times taller and correspondingly narrower, so the genuine signal is no longer drowned in numerical diffusion. The practical pay-off: when you *do* switch real diffusion on, the cohort widths you measure reflect the biology you put in rather than the grid you happened to choose.

One **caveat** worth repeating: because bin-averaging corrects an `O(Δw)` bias, turning it on will shift your reported biomass, yield and resource spectrum slightly (more on coarse grids). A model that was carefully calibrated under the first-order scheme may therefore need recalibrating. That is precisely why the feature is opt-in and flagged experimental. The new ["Numerical Details" vignette](https://sizespectrum.org/mizer/articles/numerical_details.html) and the [analytic test vignette](https://sizespectrum.org/mizer/articles/analytic_test.html) go through the mathematics and the convergence tests in full.

## A new L-stable time-stepper

On the time axis, mizer 3.0.0 already offered a second-order `"predictor_corrector"` method alongside the default `"euler"`. The Crank–Nicolson corrector it uses, however, can ring at large time steps. mizer 3.1.0 adds `method = "tr_bdf2"`, an L-stable TR-BDF2 scheme that keeps the second-order accuracy while damping those oscillations — and, like every mizer time-stepper, still only needs tridiagonal solves.

``` r
sim <- project(params, t_max = 100, method = "tr_bdf2")
```

What does "ringing" look like, and why does L-stability cure it? The Crank–Nicolson corrector is only *neutrally* stable for stiff modes: instead of damping the sharp, high-frequency features it cannot resolve at a large step, it reflects them back as oscillations. We can provoke it with the same travelling pulse as above (reusing the exact solution `N_analytic()` and growth rate `start_growth()` defined earlier) but taking a deliberately large step, `dt = 0.24`.

```{r}
#| label: cn-ringing
#| message: false
#| code-fold: true
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "At a large time step the Crank–Nicolson corrector of `predictor_corrector` rings, throwing off spurious oscillations into negative densities near the sharp recruitment boundary, while the L-stable `tr_bdf2` stays glued to the exact solution."
# Build the analytic-test pulse model (rates from the earlier chunk).
start_mort <- function(params, ...) matrix(B * params@w^(p - 1), nrow = 1)

pr <- newMultispeciesParams(data.frame(species = "Test", w_max = 1000,
                                        w_mat = 100),
                            no_w = 1000, min_w = 1e-3, info_level = 0)
pr <- setRateFunction(pr, "EGrowth", "start_growth")
pr <- setRateFunction(pr, "Mort", "start_mort")
ext_diffusion(pr)[1, ] <- K * w(pr)^(p + 1)
pr@species_params$constant_reproduction <- 0
pr <- setRateFunction(pr, "RDD", "constantRDD")

wv <- w(pr)
initialN(pr) <- matrix(N_analytic(wv, t_start, w0, 0), nrow = 1)

elapsed <- 6.4
dt_big  <- 0.24                          # a deliberately large step
final_n <- function(method)
    finalN(project(pr, t_max = elapsed, dt = dt_big, t_save = elapsed,
                   method = method, progress_bar = FALSE))[1, ]

n_pc <- final_n("predictor_corrector")
n_tb <- final_n("tr_bdf2")
n_ex <- N_analytic(wv, t_start + elapsed, w0, 0)

plot(wv, n_pc, type = "l", log = "x", lwd = 2, col = "#D55E00",
     ylim = range(n_pc, n_tb), xlim = c(1e-2, 1e3),
     xlab = "Weight [g]", ylab = "Density  N(w)",
     main = sprintf("Large time step (dt = %.2f)", dt_big))
abline(h = 0, col = "grey70")
lines(wv, n_tb, lwd = 2, col = "#0072B2")
lines(wv, n_ex, lwd = 2, lty = 2, col = "grey30")
legend("topright", bty = "n", lwd = 2, lty = c(1, 1, 2),
       col = c("#D55E00", "#0072B2", "grey30"),
       legend = c("predictor_corrector", "tr_bdf2", "exact"))
```

The true solution is the small bump on the right, where the pulse has grown to. `predictor_corrector` reproduces that bump but also spews oscillations an order of magnitude larger across the small-size end, swinging the density well below zero — unphysical, and a recipe for instability if fed back through the nonlinear rates. `tr_bdf2` damps exactly those modes and tracks the exact solution everywhere. (Both methods are equally accurate once the step is small enough; the difference is purely a large-step robustness one, explored further in the ["Numerical Details" vignette](https://sizespectrum.org/mizer/articles/numerical_details.html).)

Under both second-order time methods the resource and any extra ecosystem components are now advanced to second-order accuracy in time as well, rather than being left at first order.

## Clarifying the maximum-size parameters

mizer has long carried three parameters that all sound like "the maximum size", and over the years they have meant subtly different things. mizer 3.1.0 settles their roles. To explain why the dust kept getting unsettled, it helps to recall the history (see [issue #325](https://github.com/sizespectrum/mizer/issues/325)):

- In **mizer 1.0** there was a single maximum-size parameter, `w_inf`. It was both the size at which growth stopped *and* the size beyond which no fish could exist, and people were encouraged to set it to the von Bertalanffy asymptotic size.
- In **mizer 2.0** we noticed that real populations contain fish larger than the von Bertalanffy `w_inf`, so `w_max` was introduced as the true ceiling — the size at which growth stops and beyond which there are no fish. Existing models had `w_inf` renamed to `w_max`, and von Bertalanffy curves were discouraged.
- **Late in the 2.x series** we realised that growth *diffusion* lets some individuals push past the size at which the deterministic growth rate hits zero. So `w_repro_max` was introduced for "the size at which mature fish invest everything in reproduction", and `w_max` started to retreat into being a purely computational boundary.
- In **mizer 3.0** diffusion became fully integrated, von Bertalanffy growth rates returned as the standard way to constrain growth, and we became quite comfortable with fish living beyond the von Bertalanffy asymptotic size — because diffusion is exactly how they get there.

That history left `w_max` doing far too many unrelated jobs and `w_inf` demoted to a plotting aid. mizer 3.1.0 puts each parameter back in its place:

- **`w_inf`**, the von Bertalanffy asymptotic size, is once again the first-class, required maximum-size parameter, and supplies the default for `w_repro_max` and `w_mat`.
- **`w_repro_max`** is *only* the size at which a typical mature individual invests all its energy in reproduction. It is **not** a hard ceiling on size — a small proportion of fish may still be immature there, and diffusion can carry individuals beyond it. There is no expectation that it is larger or smaller than `w_inf`, since von Bertalanffy estimates are often unreliable.
- **`w_max`** is now *purely* a computational boundary — the upper end of the size grid and of plot ranges. When you do not supply it, it defaults to `1.5 * w_inf`.

This is a **breaking change** in spirit, but a gentle one in practice. For backwards compatibility, if you do not supply `w_inf` it is taken from `w_repro_max` or `w_max`, so **existing models and scripts continue to work unchanged**. The difference shows up only when you build a *new* model and rely on the defaults.

## Other improvements

- **Extension packages can now version and upgrade their own data** inside saved model objects, independently of the mizer version, via `recordExtension()` and an `upgrade()` S3 method. The core mizer upgrade is now itself just the `upgrade.MizerParams()` / `upgrade.MizerSim()` method.
- **`getDiet()` gains a `MizerSim` method** and `plotDiet()` accepts a `time_range`, so you can look at diet composition from the *simulated* abundances at any time rather than only the initial state (#357).
- **`getDiffusion()` now supports a fully size-dependent predation kernel**, not just a predator/prey ratio kernel (#373).
- **`getTrophicLevel()` gives the resource a size-dependent trophic level** instead of treating it as level 0, controlled by new `w_R` and `beta_R` arguments.
- **Resource functions return classed objects** with the same convenient `print()`, `summary()`, `plot()` and `as.data.frame()` methods as the consumer rate functions, so e.g. `plot(NResource(sim))` just works.
- New `getFluxGradient()` function returns the flux divergence that appears as the transport term in the mizer size-spectrum equation.
- **`summary()` of a `MizerSim`** now reports the fishing effort actually used during the run, not the model's `initial_effort`.
- A **new vignette explains how mizer calculates its default parameter values** (#189).

## Bug fixes

- `project()` was advancing extra ecosystem components only once per *saved* time step instead of once per `dt`.
- `getRDI()`, `getRDD()` and `getFlux()` on a `MizerSim` now use the simulated time-varying effort rather than the initial effort (#370).
- `plotCDF()` / `plotlyCDF()` place each cumulative value correctly on its bin's upper edge instead of its lower edge.
- `distanceMaxRelRDI()` now returns `Inf` instead of `NaN` when a previous RDI is zero, so `projectToSteady()` no longer mistakes a `NaN` distance for convergence.
- `plotYield()` now works correctly when called via `do.call()`.

## Upgrading

Install the new version from CRAN (or [GitHub](https://github.com/sizespectrum/mizer) for the development version) as usual. Models and simulations saved with older versions are upgraded automatically the first time they are validated, so there is nothing extra to do. As always, the full list of changes is in the [NEWS file](https://sizespectrum.org/mizer/news/index.html).

Happy modelling!
