---
title: "Density-dependence in reproduction"
description: |
  The take-home message is that it is important to set the density
  dependence in your model appropriately and that with the help of the
  `setBevertonHolt()` function you can easily make changes to the density
  dependence without spoiling other aspects of your calibrated model.
author:
  - name: Gustav Delius
    url: {}
date: 08-03-2021
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this blog post I want to discuss the purpose of the `setBevertonHolt()`
function, which was improved in the latest 
[version 2.2.1](https://sizespectrum.org/mizer/news/index.html) 
of mizer to make it convenient to control the density-dependence in
reproduction. We'll introduce the new concept of the reproduction level and meet
the related new function `getReproductionLevel()`.

Before we get started we need to clarify what we mean by density dependence in
reproduction in a mizer model.

# Stock-recruitment relationship

Usually, in a fisheries model, density dependence is specified via a
stock-recruitment relationship, which gives a non-linear relationship between
the spawning stock biomass and the recruitment. This relationship could, for
example, be described by a Beverton-Holt curve. In that case for low spawning
stock biomass the recruitment is assumed to be linearly proportional to the
spawning stock biomass but for higher spawning stock biomass the curve flattens
and asymptotically approaches a maximum recruitment rate that will never be
exceeded, no matter how high the spawning stock biomass.

In a mizer model things are different because many of the causes of the density
dependence in recruitment are explicitly modelled mechanistically, so don't need
to be imposed externally. These causes are:

1. The rate at which a spawner invests energy into reproduction depends on
its food intake. So if there is a high abundance of fish and hence a lot
of competition for prey, each spawner will produce fewer offspring.

2. Fish larvae are exposed to  mortality from predation by larger fish. So
the higher the abundance of larger fish, the higher the mortality and the
fewer of the offspring will reach recruitment size.

3. The rate at which fish larvae grow, and hence the time they take to reach
recruitment size, will depend on the availability of resources. If there are
many fish larvae competing for limited resources, they will be exposed to
mortality for longer, meaning fewer of them will reach recruitment size.

All these causes of density dependence are automatically taken into account in
mizer, because growth depends on the abundance of prey and mortality depends on
the abundance of predators. This dynamically generates fully multi-species
stock-recruitment relationships.

However, there are other causes of density dependence besides the three listed
above, and these are not explicitly modelled mechanistically by mizer. An
example would be the limited carrying capacity of suitable spawning grounds and
other spatial effects. The way such additional density dependencies can be taken
into account in a mizer model is by specifying a non-linear relationship between
the energy that the spawners invest into reproduction and the number of larvae
this results in. It is this extra density-dependence that we will be concerned
with in this blog post.

# An example model

To make things concrete in this blog post we will use the example 
MizerParams object `NS_params`, included in mizer, that describes the fish
community in the North Sea. Let's have a quick look at that MizerParams
object.

First we load the mizer package. We are using the currently latest version
of mizer, version 2.2.1.

```{r message=FALSE}
library(mizer)
packageVersion("mizer")
```

A MizerParams object comes with initial size spectra for all species and the
resource. Let's plot these for the North Sea params object.

```{r}
params <- NS_params
plotSpectra(params, power = 2)
```

These initial spectra were chosen to describe a steady state of the model,
meaning that if we project forward in time with constant fishing effort, the
abundances stay constant. Let's verify that.

```{r}
sim <- project(NS_params, t_max = 20)
plotBiomass(sim)
```

We'll use this model below

# Beverton-Holt curves

By default, mizer uses a Beverton-Holt curve to describe the relationship
between the rate $E_R$ at which energy is invested into reproduction and the
rate $R$ at which offspring are produced. 

To make this concrete, we'll look at an example. I have written a function for
the purpose of this blog post called `plotBevertonHolt()` which we can use to
look at the Beverton-Holt curve for a given species in a model.

```{r warning=FALSE, code_folding=TRUE}
library(ggplot2)
plotBevertonHolt <- function(params, species) {
  select <- species_params(params)$species == species
  erepro <- species_params(params)$erepro[select]
  w0 <- params@w[params@w_min_idx[select]]
  E_R_ss <- getRDI(params)[select] / erepro * 2 * w0
  R_dd_ss <- getRDD(params)[select]
  R_max  <- species_params(params)$R_max[select]
  E_R <- seq(0, 2 * E_R_ss, length.out = 50)
  R_di = erepro * E_R / 2 / w0
  R_dd <- R_di / (1 + R_di / R_max)
  df <- melt(data.frame(E_R, R_dd, R_di, R_max), id.vars = "E_R")
  ggplot(df) +
    geom_line(aes(x = E_R, y = value, linetype = variable)) +
    geom_point(aes(x = E_R_ss, y = R_dd_ss), size = 3, color = "red") +
    ylim(NA, 1.1 * R_max) +
    ylab("Reproduction rate [eggs/year]") +
    xlab("Energy invested [g/year]")
}
```

We use that function to plot the Beverton-Holt curve for Gurnard in our
example model

```{r warning=FALSE}
plotBevertonHolt(params, "Gurnard")
```

The solid line is the Beverton-Holt curve. The horizontal dashed line shows the
maximum reproduction rate that the Beverton-Holt curve approaches for large
$E_R$ and the dotted line shows the density-independent reproduction.
The current reproduction rate is marked on the curve by a red dot. The curve
shows us how the reproduction rate would change as the rate of investment into
reproduction changes.

There are two parameters involved in determining this curve:

1. The reproductive efficiency `erepro` determines the slope of the
curve when reproduction is low,

2. The maximum reproduction rate `R_max` determines the asymptote to which
the curve levels off eventually.

Let us look up the values to which  the reproductive efficiency `erepro` 
and the maximal recruitment rate `R_max` are set for Gurnard in this
example model.

```{r}
select_gurnard <- species_params(params)$species == "Gurnard"
species_params(params)[select_gurnard, c("erepro", "R_max")]
```


# The reproduction level

We define the __reproduction level__ as the ratio between the current reproduction
rate and the maximum reproduction rate. Thus it is a number between 0 and 1.
It is similar to the __feeding level__ which in mizer is defined as the ratio
between tha actual feeding rate and the maximum feeding rate. 

Since version 2.2.1 of mizer there is a function `getReproductionLevel()` for
calculating the reproduction level.

```{r}
reproduction_level <- getReproductionLevel(params)
reproduction_level
```

We see that the reproduction level of Gurnard is
`r reproduction_level[select_gurnard]`, which agrees with the height of the red
dot in relation to the maximum in the above plot of the Beverton Holt
curve for Gurnard. 

Most of the other species have unrealistically large reproduction levels, very
close to 1. To see why that is bad let's look for example at the curve for Cod:

```{r warning=FALSE}
plotBevertonHolt(params, "Cod")
```

This looks so strange because the actual reproduction rate is very close to the
maximum reproduction rate, so that we are very far to the right of the
Beverton-Holt curve where it is almost flat. The density dependence here is so
strong that an increase or decrease in the energy invested into reproduction
leads to almost no change in the reproduction. In other words, the spawning
stock biomass has almost no effect on the number of eggs produced. That is not
realistic, so before using this model we should reduce the reproduction level.

# Changing the reproduction level

We could reduce the reproduction level by either reducing the reproductive
efficiency `erepro` or by increasing `R_max`, but there is a subtlety.
Given that the reproduction level is the ratio of the actual reproduction rate
and the maximum reproduction rate `R_max`, we might think that doubling
`R_max` would reduce the reproduction level by a factor of 2. Let's check
that in the example of Cod:

```{r}
select_cod <- species_params(params)$species == "Cod"
params_changed_R_max <- params
species_params(params_changed_R_max)$R_max[select_cod] <- 
    2 * species_params(params_changed_R_max)$R_max[select_cod]
getReproductionLevel(params_changed_R_max)[select_cod]
```

The reproduction level has changed very little. The reason is that besides
raising the maximum reproduction, we have also changed the Beverton-Holt
curve and the actual reproduction has also increased. 

The other problem with changing either `erepro` or `R_max` in isolation is
that we will move away from the steady state. We can see that by projecting
our changed params object forward in time:

```{r}
sim_changed_R_max <- project(params_changed_R_max, t_max = 20)
plotBiomass(sim_changed_R_max, species = "Cod")
```

This is bad news if we had carefully calibrated our model to observed
biomasses for example.

What we need to do is change both `erepro` and `R_max` at the same time in
such a way that the actual reproduction does not change. This is what the
function `setBevertonHolt()` does.

```{r}
params2 <- setBevertonHolt(params, reproduction_level = c(Cod = 0.5))
plotBevertonHolt(params2, "Cod")
```

To see more clearly what has happened, we extend our `plotBevertonHolt()`
function to show both the old and the new Beverton-Holt curve in the
same graph.

```{r, code_folding=TRUE}
plotBevertonHolt2 <- function(params, params2, species) {
  select <- species_params(params)$species == species
  erepro <- species_params(params)$erepro[select]
  w0 <- params@w[params@w_min_idx[select]]
  E_R_ss <- getRDI(params)[select] / erepro * 2 * w0
  R_dd_ss <- getRDD(params)[select]
  E_R <- seq(0, 2 * E_R_ss, length.out = 50)
  
  R_max  <- species_params(params)$R_max[select]
  R_di = erepro * E_R / 2 / w0
  R_dd <- R_di / (1 + R_di / R_max)
  df <- melt(data.frame(E_R, R_dd, R_di, R_max), id.vars = "E_R")
  df$Model <- "Model 1"
  
  erepro <- species_params(params2)$erepro[select]
  R_max  <- species_params(params2)$R_max[select]
  R_di = erepro * E_R / 2 / w0
  R_dd <- R_di / (1 + R_di / R_max)
  df2 <- melt(data.frame(E_R, R_dd, R_di, R_max), id.vars = "E_R")
  df2$Model <- "Model 2"
  
  ggplot(rbind(df, df2)) +
    geom_line(aes(x = E_R, y = value, linetype = variable,
                  colour = Model, size = Model)) +
    geom_point(aes(x = E_R_ss, y = R_dd_ss), size = 3, color = "red") +
    ylim(NA, 1.1 * R_max) +
    ylab("Reproduction rate [eggs/year]") +
    xlab("Energy invested [g/year]") +
    labs(linetype = "", size = "R_max", colour = "R_max") +
    scale_size_manual(values = c(0.5, 1)) +
    scale_colour_manual(values = c("blue", "black")) +
    scale_linetype_manual(values = c("solid", "dashed", "dotted"))
}
```

```{r, preview=TRUE}
plotBevertonHolt2(params, params2, "Cod")
```

This shows that the red dot that marks the steady state value lies on both
the old (blue) and the new (black) curve. 
This was achieved by lowering `erepro` at the same
time as increasing `R_max`. The old values were

```{r}
species_params(params)[select_cod, c("erepro", "R_max")]
```

and the new values are

```{r}
species_params(params2)[select_cod, c("erepro", "R_max")]
```

So the change in the reproduction level has been
achieved without a change to the steady state. We can verify this:

```{r}
sim2 <- project(params2, t_max = 20)
plotBiomass(sim2)
```

# Sensitivity to fishing

The reason why it matters what level of density dependence in reproduction
is chosen for the model is that it affects the sensitivity of the model to
perturbations, for example to changes in fishing pressure. To illustrate this
we will plot the sustainable fishing yield as a function of fishing mortality.

The `mizerExperimental` package contains a function `plotYieldVsF()` that 
creates such a Yield versus F plot. The `mizerExperimental` package is 
updated frequently and thus it is a good idea to install the latest version

```{r message=FALSE, warning=FALSE, eval = FALSE}
remotes::install_github("sizespectrum/mizerExperimental")
```


```{r}
library(mizerExperimental)
packageVersion("mizerExperimental")
```

Here is the Yield versus F plot for Cod in the old model with the high
reproduction level:

```{r message=FALSE}
plotYieldVsF(params, "Cod")
```

This plot shows the _sustainable_ fishing yields at different levels of imposed
fishing mortality. This is calculated by running the model with each level of
fishing mortalit for long enough to reach a steady state. Then the yield in that
steady state is plotted. For high fishing mortality the stock should collapse
and the yield in the steady state should therefore be very low at high fishing
mortalities. However in this model we see that the stock can sustain
unrealistically high fishing mortalities without collapsing.

For the new model, with a reproduction level of 1/2, the curve looks more
realistic:

```{r}
plotYieldVsF(params2, "Cod")
```

In this model at a fishing mortality of 2 per year the stock collapses 
completely, leading to zero yield.

# More about `setBevertonHolt()`

As you can see from the 
[help page](https://sizespectrum.org/mizer/reference/setBevertonHolt.html) 
of `setBevertonHolt()`, you can adjust the density dependence in the
reproduction of several species at once. For example we can try to set the
reproduction level to 1/2 for all species:

```{r warning=TRUE}
params3 <- setBevertonHolt(params, reproduction_level = 0.5)
```

This gave us a warning. Let's take a look at plaice, for example:

```{r}
select_plaice <- species_params(params)$species == "Plaice"
species_params(params3)[select_plaice, c("erepro", "R_max")]
```

We see that to achieve a reproduction level of 1/2 while maintaining plaice at
its initial abundance, it would need a very high reproductive efficiency.  

Let's see what happens if we try to force the reproductive efficiency to a more
realistic level of 0.1:

```{r warning=TRUE}
params3 <- setBevertonHolt(params, erepro = c("Plaice" = 0.1))
```

```{r}
species_params(params3)[select_plaice, c("erepro", "R_max")]
```

To understand what is going on here we have to recall that `setBevertonHolt()`
will not make changes that would change the abundances in the model.
The level of abundance for plaice specified in the model can only be achieved with a high reproductive efficiency. 

If we want to force the change in
`erepro` we can do that but have to accept that the steady state will change: 

```{r}
species_params(params3)$erepro[select_plaice] <- 0.1
sim3 <- project(params3, t_max = 20)
plotlyBiomass(sim3)
```

We see that this change also affects the abundances of the other species. 
I think this nicely illustrates the differences of adjusting reproduction
parameters with `setBevertonHolt()` as opposed to changing them directly.

# Turning off density dependence

We can also set the reproduction level to 0.

```{r}
params4 <- setBevertonHolt(params, reproduction_level = 0)
```

This does not mean that we have set reproduction to 0, but rather that we have
turned off the density dependence in reproduction. In other words, the
reproduction level is zero because we have set the maximum reproduction rate to
infinity.

```{r}
species_params(params4)$R_max
```

I want to stress that the fact that we have turned off all the density 
dependence in reproduction does not mean that the stock-recruitment 
relationship is linear. All the density dependencies mentioned at the start of
this blog post are still operating. We are simply not imposing any additional
density dependence that is not already captured by the model. 

The built-in density dependencies are enough to keep the model stable, in the
sense that if we make changes to the initial abundances, after a time the system
will settle down to a steady state again:

```{r}
params5 <- params4
initialN(params5) <- 4 * initialN(params4)
sim5 <- project(params5)
plotlyBiomass(sim5)
```

# Summary
We have discussed that in the mizer model many sources of density dependence
in recruitment are captured automatically but that one can take into account
additional density dependence in recruitment by imposing a density dependence
on the reproduction, by which we mean to impose a non-linear relation between
the rate at which energy is invested into reproduction and the rate at which
offspring are produced.

We looked at some Beverton-Holt curves which are used in mizer by default to
encode this non-linearity. These curves are specified by two parameters:
the reproductive efficiency `erepro` and the maximum reproduction rate `R_max`.
The `setBevertonHolt()` function allows us to change both of these parameters
together without changing the current reproduction rate. It only affects the
behaviour of the model when it is perturbed away from the current state.

Therefore the `setBevertonHolt()` function allows us to control the sensitivity
of the model to perturbations, like changes in fishing for example, without
changing the steady-state properties. This allows a two-stage approach to
calibrating a mizer model: in a first step one calibrates the steady state
properties to reproduce averaged observations, like growth rates and biomasses
or yields of species for example. In a second step on the calibrates the
sensitivity of the model, for example by matching it to time series
observations.