---
title: "A 5-step recipe for tuning the model steady state"
description: |
  Getting a steady state for your mizer model that agrees with observations
  is in principle a hard chicken and egg problem. I present the trick that
  makes it surprisingly easy, with a 5-step recipe. I'll save tips on what
  to do when the recipe fails for later blog posts.
author:
  - name: Gustav Delius
    url: {}
date: 08-20-2021
twitter:
  site: "@mizer_model"
  creator: "@gustavdelius"
output:
  distill::distill_article:
    self_contained: false
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# What we want to do

In this blog post we will describe stage 2 of the process of building a mizer
model. The stages are:

1. Collect information about the important species in your ecosystem and how
they are fished. This includes physiological parameters for the species as
they might be found on fishbase, but also information about how abundant the
species are and how they are being fished. 

2. Create a mizer model that in its steady state  reproduces the time-averaged
observed state of your fish community. Of course your real system never is in a
perfect steady state. It is continuously changing. There is much fluctuation
from year to year. We will however assume that if we average observations over a
number of years we obtain something that is close to the steady state. Without
some such assumption it would be impossible for us to get started.

3. Tune the model parameters further to also
reproduce time-series observations that capture some of the system's sensitivity
to perturbations, like changes in fishing pressure. 

This blog post is only about the second stage. We will present a 5 step recipe
for that stage. When the recipe works, it will only take a couple of minutes! So
I hope you will try the recipe for your own model. Of course in practice there
are all kinds of things that can (and will) go wrong. So I hope this blog
post will lead to some exchange of experiences with the recipe.

The recipe is based on an important trick. I call it the constant reproduction
trick. The trick is obvious once you see it, but I must admit that I struggled
for a long time with mizer model building until I stumbled upon the trick.

# Example we will use

To make this concrete, we will consider a model for the North Sea involving
12 species. We will short-circuit stage 1 of the model-building process by
basing our example on the North Sea species parameter data frame
`NS_species_params` that comes with mizer.

Some of the functions we will be using are still in active development in the
[mizerExperimental](https://sizespectrum.org/mizerExperimental/) package. 
Therefore we will always want to make sure we are loading the latest version of
the package with

```{r}
remotes::install_github("sizespectrum/mizerExperimental")
library(mizerExperimental)
```

This blog post was compiled with mizer version  `r packageVersion("mizer")`
and mizerExperimental version `r packageVersion("mizerExperimental")`

Here is the species parameter data frame that we will be using:

```{r, code_folding=TRUE, layout="l-body-outset"}
# Here is how I obtained the example species_params:
species_params <- NS_species_params
species_params$R_max <- NULL
species_params$a <- c(0.007, 0.001, 0.009, 0.002, 0.010, 0.006, 0.008, 0.004,
                         0.007, 0.005, 0.005, 0.007)
species_params$b <- c(3.014, 3.320, 2.941, 3.429, 2.986, 3.080, 3.019, 3.198,
                         3.101, 3.160, 3.173, 3.075)

years <- getTimes(NS_sim) >= 1990 & getTimes(NS_sim) <= 2010
# Average biomass above w_mat/20
bm_hist <- getBiomass(NS_sim, min_w = species_params$w_mat / 20)[years, ]
species_params$biomass_observed <-  colSums(bm_hist) / 21

library(knitr)
kable(species_params, row.names = FALSE)
```

For each species we are specifying its name and some parameters
characteristic of the species: its asymptotic size `w_inf` and maturity size
`w_mat`, the parameters `beta` and `sigma` for its feeding kernel (we are using
the default lognormal kernel for all species), the von Bertalanffy growth
parameter `k_vb` and the parameters `a` and `b` in the allometric length-weight
relationship $w = a l^b$. 

In addition, we also specify information that is specific to our ecosystem,
namely the average abundance of each species, in the `biomass_observed` column.
This is measured in grams. Because for the purpose of this blog post it is not
important, we did not bother to look up real biomass estimates but instead we
simply used the average over the years 1990 to 2010 in the simulated data in the
`NS_sim` object included in the mizer package. You are invited to re-run the
analysis with proper data.

The observed system is being fished. We need to give mizer information about
how it is being fished. We do this via the gear_params data frame.

```{r, code_folding=TRUE, layout="l-body-outset"}
# Average fishing mortality
f_location <- system.file("extdata", "NS_f_history.csv", package = "mizer")
f_history <- as(read.csv(f_location, row.names = 1), "matrix")[years, ]
f <- colSums(f_history) / 12

gear_params <- 
    data.frame(gear = "All",
               species = NS_species_params$species,
               sel_func = "sigmoid_length",
               l25 =  c(7.6, 9.8, 8.7, 10.1, 11.5, 19.8, 16.4, 19.8, 11.5,
                        19.1, 13.2, 35.3),
               l50 = c(8.1, 11.8, 12.2, 20.8, 17.0, 29.0, 25.8, 29.0, 17.0,
                       24.3, 22.9, 43.6),
               catchability = f)

kable(gear_params, row.names = FALSE)
```

We are setting up a single gear that we call "All" which catches all species.
For each species we set up the selectivity curve of the gear as a sigmoid curve
with given `l25` and `l50` parameters. Finally we set the catchability of each
species to the observed fishing mortality, averaged over the years 1990 to 2010.
We will then set the fishing effort to 1, because in mizer the fishing mortality
is the product of effort, catchability and selectivity.

Our task now is to create a mizer model that describes species with the above
characteristics and that has a steady state with the observed biomasses under
the given fishing pressure.

# Why it is a hard problem

We have a chicken and egg problem. The equilibrium abundance and size
distributions of the fish are determined by their size-dependent growth and
death rates. These rates in turn are determined by the abundance and size
distribution of their prey and their predators. So we can't determine the size
distributions before we have determined the rates and we can't determine the
rates before we have determined the size distributions.

Because every species is a both prey and predator of fish of various species and
sizes during their life, this is a highly coupled non-linear problem. If for
example we use 100 size bins and 12 species, plus a resource spectrum, then we
would have far over a thousand coupled nonlinear equations to solve
simultaneously. That is not practical.

Rather than solving the equilibrium equations, another way to find a steady
state is to simply evolve the time dynamics until the system settles down to a
steady state. The problem with this approach is that the coexistence steady
state of a size spectrum model has a very small region of attraction, so unless
one starts with an initial state that is already very close to that coexistence
steady state one will end up with extinctions.

The reason is a feedback loop: as the spawning stock biomass of a species grows,
also its reproduction rate grows, leading to further growth of the spawning
stock biomass and so on. Similarly as the spawning stock of another species
declines, so does its reproduction rate, leading to further decline. In spite of
moderating non-linear effects in the model, the general outcome is extinctions.

We can see the phenomenon in our North Sea example. If we simply run the
dynamics, starting with the intial state set up by `newMultispeciesParams()`,
first Sprat goes extinct, and Herring follows soon after. Just click the play
button on the animation below.

```{r}
params <- newMultispeciesParams(species_params = species_params,
                                gear_params = gear_params,
                                initial_effort = 1)
sim <- project(params, t_max = 12)
animateSpectra(sim, power = 2)
```


# The constant reproduction trick

So the trick is to cut the destabilising feedback loop by decoupling the
reproductive rate from the spawning stock biomass. We do this by simply keeping
the reproduction rate constant. The size spectrum model with constant
reproduction turns out to be very stable and quickly approach a steady state,
due to the smoothing effect of the feeding kernel. Once the steady state is
found, we can simply adjust the reproductive efficiency of each species so that
the steady state spawning stock produces the chosen reproduction rate. With that
choice of the reproductive efficiency the steady state of the restricted
dynamics is also the steady state of the full size spectrum model.

Here is the code that does that. Run the animation that it produces by 
clicking the play button.

```{r}
params <- newMultispeciesParams(species_params = species_params,
                                gear_params = gear_params,
                                initial_effort = 1)
# Keep reproduction constant at the initial level
params@species_params$constant_reproduction <- getRDD(params)
params <- setReproduction(params, RDD = "constantRDD")
# Run the dynamics with this constant reproduction
sim <- project(params, t_max = 15)
animateSpectra(sim, power = 2)
```

Mizer has a function called `steady()` that does the same as the above code,
namely run to steady state with constant reproduction and then adjust the
reproduction parameters, and then sets the resulting steady state as the 
initial state of the MizerParams object.

```{r}
params <- newMultispeciesParams(species_params = species_params,
                                gear_params = gear_params,
                                initial_effort = 1)
params <- steady(params)
plotlySpectra(params, power = 2)
```

We now have a MizerParams object whose initial state is a steady state. 
Running a simulation starting with these initial conditions will show no
change over time. For example the biomasses of all species will stay 
constant. 

```{r}
sim <- project(params, t_max = 5)
plotBiomass(sim)
```

But how do these biomasses compare to our observed biomasses?

```{r}
plotBiomassVsSpecies(params)
```
They don't agree at all, but that is no surprise. It would actually have been 
quite a coincidence if they did agree, because the `newMultispeciesParams()`
function did not know how big our ecosystem is. It did not know that we wanted
the biomasses in the entire North Sea. So initially the scale is arbitrary.
The dynamics of the model are obviously independent of the scale of the system.
So we have the freedom to change that scale. The `calibrateBiomass()`
function chooses the scale so that the total biomass in the model agrees
with the total observed biomass.

```{r, preview=TRUE}
params <- calibrateBiomass(params) 
plotBiomassVsSpecies(params)
```
So now the total biomass is correct, but for some species the biomass in the
model is too high, for others it is too low. 

Actually, that the size spectrum is too low for Saithe and Cod and too high for
Dab and Haddock might also be suspected from the fact that they are outliers in
the size-spectrum plot above. We expect in a healthy ecosystem that the total
spectrum roughly follows a power law, i.e., a straight line on the log-log plot.
Those species currently spoil that.

So we want to lower the spectra of the species whose biomass is too high in the
model and raise those of the species whose biomass is too low. This is what
the `matchBiomasses()` function does.

```{r}
params <- matchBiomasses(params)
plotlySpectra(params, power = 2, total = TRUE)
```

In fact, it has raised and lowered the spectra by exactly the factor needed
to get the model biomasses to match the observed biomasses.

```{r}
plotBiomassVsSpecies(params)
```

Of course this is not the end of the story, because just rescaling the size
spectra by constants will not again produce a steady state. All species now
experience a new prey distribution and a new predator distribution, so their
growth and death rates have changed. We will have to again run the dynamics to
steady state.

```{r}
params <- steady(params)
plotBiomassVsSpecies(params)
```

This has now spoiled the agreement between observed and model biomasses. But
we can simply calibrate and match again and run to steady state again.

```{r}
params <- params |> calibrateBiomass() |> matchBiomasses() |> steady()
plotBiomassVsSpecies(params)
```

The discrepancies are now quite small. We could iterate to get them even
smaller:

```{r}
params <- params |> calibrateBiomass() |> matchBiomasses() |> steady() |>
    calibrateBiomass() |> matchBiomasses() |> steady()
plotBiomassVsSpecies(params)
```

So here is the picture of the steady state that matches the observed biomasses:

```{r}
plotlySpectra(params, power = 2)
```

Actually, even the growth rates in the steady state match the von Bertalanffy
growth curves pretty well:

```{r}
plotGrowthCurves(params, species_panel = TRUE)
```

But this is a bit of a coincidence. Mizer has to choose values for the
coefficient `gamma` of the 'search volume' for each species and for the
coefficient 'h' of the maximum intake rate, both of which affect the growth
rates. Because mizer has to choose them before it knows what the steady state
prey distribution is for each species, it can not guarantee to choose them so as
to give the correct growth rates in the steady state. You usually will have to
retune them by hand. The mizerExperimental package provides a convenient shiny
gadget that allows you to do that interactively, and I will talk about that in
future blog posts.

Also remember that getting the steady state to agree with time-averaged
observations is just the second stage in tuning a mizer model. Next you will
want to tune the sensitivity to changes away from steady state. This will in
particular involve tuning the 
[density dependence in reproduction](https://blog.mizer.sizespectrum.org/posts/2021-08-03-density-dependence-in-reproduction/), 
among other things.


# Summary of the recipe

We have seen how to proceed if you have your species parameters and gear
parameters and also have averaged observed biomasses for each species that you
want the steady state of your model to match:

1. Create a MizerParams object from your species parameters and gear parameters
with `newMultispeciesParams()`.

2. Find a coexistence steady state with `steady()`.

3. Set the scale of the model to agree with the observed total biomass with
`calibrateBiomass()`. This does not spoil the steady state.

4. Use `matchBiomass()` to move the size spectra of the species up or down
to match the observed biomasses. This will spoil the steady state.

5. Go back to step to to again find the steady state. Iterate steps 2, 3 and 4
as often as you like to get the steady-state biomasses to agree as precisely
with your observations as you like.

There are several interesting ways in which the above recipe can fail.
I'll blog about them in the future. But it will be more fun if you share your
attempt at following the above recipe with your species parameters and your
observed biomasses. Email me at <gustav.delius@gmail.com>. I can then use your
example to explain what to do when problems arise.

